Supp mat: time series paper

### 2017-07-07 writing/manuscript_draft.tex
updated citations and spelling throughout document

To be submitted to: \emph{Biological Conservation} (Short Communication)
\vspace{3 mm}

@@

Long-term census data is necessary to better understand ecological
processes and make management decisions. However, census data is often
expensive, requiring a lot of people-hours and equipment. Little work
has addressed the question, when is a time series of census data long
enough to address a question of interest? Here, we explore two
approaches to address this question: through simulations and through an
empirical approach. We specifically determine the minimum time series
length required to estimate significant increases or decreases in
population abundance. Importantly, We examine the ability to detect a
trend with a set level of statistical power, which is often neglected in
ecological time series analyses. Using a simulation approach, we
demonstrate how the minimum time series length required increases with
weaker trends in abundance and with higher population variability. In
addition, we examine 868 populations of vertebrate species to determine
the minimum time required to detect changes in their abundance. We show
that 10-15 years of continuous monitoring are required in order to
achieve a high level of statistical power. Similar to
simulation-approaches, the minimum time required for field census data
strongly depends on trend strength, population variability, and temporal
autocorrelation. These results point to the importance of sampling
populations over long periods of time. Most studies, especially those
less than 10-15 years, are probably underpowered and potentially
misleading.

@@

Observational studies and population censuses have become a cornerstone
of modern ecological research and conservation biology (Magurran et al.
2010; Hughes et al. 2017). Long-term data is necessary to both
understand population dynamics and to assess extinction risk of species.
Even though many time series may now be considered ``long-term'', most
are still short. Time series are typically short for a variety of
reasons. They are often coupled with an experiment, which may only last
a couple of years. In addition, short funding cycles make it difficult
to examine populations over longer periods of time (Hughes et al.
2017){]}. There are, of course, notable exceptions where long-term
monitoring has occured, for example the continuous plankton recorder
(Giron-Nava et al. 2017).

Naturally, one might ask, how long of a time series is actually
necessary? This question has important implications for both research
and management. Scientifically, we need to know the time series length
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when a trend in abundance over time is actually meaningful or
not. The IUCN Red List Categories and Criteria suggest under Criterian
A2, a species qualifies as vulnerable if it has experienced a 30\%
decline over 10 years, or 3 generations (IUCN 2012). For both scientific
and management questions, because sampling is typically expensive, we do
not want to sample for longer than is neccessary (Gerber, DeMaster, and
Kareiva 1999).

@@

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (Bolker 2008). Instead of using classic statistical
tests (e.g.~t-test), and corresponding formulas for determinig sample
size, simulation approaches can be used (Gerrodette 1987, Bolker2008).
Simple models can be simulated with parameter values corresponding to a
population of interest. Out of a large number of simulations, power can
be calculated as the fraction of simulations that meet some criteria.
The specific criteria depend on the question at hand. Then, time series
simulations of different lengths are run to determine the minimum time
series length required to be confident that you will be able to answer
your question (Bolker 2008). In addition to using simulated data,
empirical time series can also be used. We usually do not have multiple
replicates of similar populations, but we can subsample an empirical
time series (Gerber, DeMaster, and Kareiva 1999). Subsamples of
different lengths can then be evaluated to see which fraction of
subsamples meet some criteria, again a measure of statistical power.
Similar to the simulation approach, we can use these measures of power
to determine the minimum time series required for a particular question
of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized amodel specifically for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organim, 15 years was adequate to assess long-term trends in
abundance. However, these studies, and other past work, have typically
focused either on theoretical aspects of monitoring design or focused on
single species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. More specifically, we estimate the minimum time
series length required to asssess long-term changes in abundance via
simple linear regression. We examine 878 population time series to
estimate this minimum time series length required.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses (P. C.
Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
approach as it has been desribed elsewhere (citations). Essentially, we
take a stochastic population model and simulate it for a number of years
repetitively. We can then determine the minimum time series required to
achieve designed levels of significance and power for a particular
question. This approach requires us to determine values for our model
parameters (e.g.~growth rate). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8). Statistical power of 0.8 would indicate that, if
there was a true trend in abundance, we would have a 0.8 probability of
detecting the trend.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) example of simulated time series for 40 time periods and
(b) statistical power versus the simulated time series length. The
horizontal, dashed line is our desired statistical power of 0.8. The
vertical, dashed line is the minimum time required to achieve our
desired statistical power. (c) Minimum time required for different
values of the trend strength (\(r\)) and (d) minimum time required for
different levels of population variability
(\(\sigma\)).\label{fig:theoretical_approach}}
\end{figure}

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The database includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. Lastly, we also calculated the minimum number of years
required (10 years or 3 generations) under IUCN Red List Criterian A2 to
classify a species as vulnerable (IUCN 2012).

For a subset of populations in the database (\(n =\) 547), we were able
to connect a database on biological characteristics (Myhrvold et al.
2015) like body size and generation length. All 547 populations are
within the Aves class. In the supplementary material, we explore how the
minimum time required does not strongly correlate with any biological
characteristics.

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 40
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then we
determined the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the long-term, or ``true``, time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1 - \beta\), where \(\beta\) is the
probability of a type II error). Lastly, we determine which subsample
length is required to guarantee we have above a certain threshold of
statistical power, which we will assume is 0.8 by convention (citation).

@@

We also examined a subset of populations where more explanatory
variables were known (e.g.~body mass, longevity, location). None of
these explanatory variables were predictive of the minimum time series
length required (\textbf{supp mat}). Some differences if \(T_{min}\) did
occur between biological class (\textbf{supp mat}), but this can be
attributed to the amount populations are known to fluctuate (i.e.~the
variability in population size).

@@

In the supplmentary material, we explore several additional questions
related to statistical power and estimating \(T_{min}\). First, we
showed that other metrics besides estimating the slope coefficient from
linear regression could be used. More specifically, we show how the same
approach desribed here can be used to determine the minimum time
required to estimate a population's geometric growth rate (Figs.
\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored
how the simulation approach can be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
More specifically, we determine the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). We found that over 27 years of continous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we
determine the minimum time required to estimate long-term trends
according to generalized additive models, instead of the simple linear
models used here (Fig. \ref{fig:gam_example}). We found that although we
obtain a similar distribution of minimum times required for GAMs, the
minimum time required for GAMs is on average shorter than for linear
regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these
results in the supplementary material.

@@

Here, we focus on an empirical approach to estimate the minimum time
series length requried to assess changes in abundance over time. We
examined long time series and then subsampled each to determine the
minimum time required to achieve a desired level of significance and
power for linear regression. Statistical power is important as it
provides on information as to the neccessary samples required to
determine a significant trend (Legg and Nagy 2006). We found that for
878 populations, at least 10-15 years of continuous monitoring were
neccessary (Fig. \ref{fig:min_time_dist}). This timeframe is in line
with past work on a short-lived snail species (Rueda-Cediel et al. 2015)
and a long-lived whale species (Gerber, DeMaster, and Kareiva 1999). In
line with theoretical predictions (Rhodes and Jonzen 2011), we also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to our prior expectations, we also found
that \(T_{min}\) did not correlate with any biological variables of
interest (Fig. \ref{fig:biological_correlates}). We initially
hypothesized that species with longer lifespans or generation times may
require a longer sampling period. Our result could have been a result of
at least two effects. First, the database we used may not include a
diverse enough set of species with different life history traits.
Second, the question we poised, whether a population is increasing or
decreasing, was specifically concerned with trends in population density
over time. Therefore, life-history characteristics may be more important
for other questions (e.g.~extinction risk) where estimates minimum time
required are needed.

@@

Our work has some limitations in determining the minimum time series
length required. First, \(T_{min}\) is particular to the specific
question of interest. For example, in the supplementary material, we
used generalized additive models to detect population trends and came up
with a similar, but quantivaely different value for \(T_{min}\) (Fig.
\ref{fig:min_time_dist_gam}). An additional complication is that for our
empirical approach, the subsampling of the full time series allows for
estimates of power, but the individual subsamples are clearly not
independent of one another. Further, estimates of \(T_{min}\) depend on
chosen values of \(\alpha\) and \(\beta\) (Fig.
\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a
specific population model parameterized for each population of interest.
Then, model simulations could be used to estimate the minimum time
series required to address each specific question of interest. Clearly,
this is not always practical, especially if conducting analyses for a
wide array of species as we do here. In addition, our statistical models
suggest that \(T_{min}\) does not correlate with any biological
variables of interest, at least for the question of linear regression
(Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to
use these results to predict \(T_{min}\) for another population, even if
the population is of a species with a similar life-history to one in our
database.

@@

We use a database of 878 populations to determine the minimum time
series length required to determine population trends. This goes beyond
previous work that either focused on theoretical investigations or a
limited number of species. We show that to identify long-term changes in
abundance, 10-15 years of continuos monitoring are required (Fig.
\ref{fig:min_time_dist}). In line with theoretical predictions
(Gerrodette 1987), we also show that \(T_{min}\) is strongly correlated
with the long-term population trend (i.e.~rate of increase), variability
in population size, and the temporal autocorrelation (Fig.
\ref{fig:correlates}). Our work implies that for many populations,
census data less than 10-15 years are probably not reliable. This
stresses the importance of long-term monitoring programs. From both a
scientific and management perspective estimates of \(T_{min}\) are
important. If a time series is too short we lack statistical power to
reliably detect long-term population trends. In addition, a time series
that is too long may be a poor use of already limited funds (Gerber,
DeMaster, and Kareiva 1999). Future work should examine other species,
with a wider range of life history characteristics. In addition, similar
approaches can be used to determine the minimum time series length
required to address additional questions of interest.

@@

@@

\hypertarget{ref-Johnson2015}{}
Johnson, Paul C.D., Sarah J.E. Barry, Heather M. Ferguson, and Pie
Müller. 2015. ``Power analysis for generalized linear mixed models in
ecology and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2):
133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

@@

\hypertarget{ref-Morrison2008}{}
Morrison, Sharn, and David S. Hik. 2008. ``When? Where? And for how
long? Census design considerations for an Alpine Lagomorph, the Collared
pika.'' In \emph{Lagomorph Biology}, 103--13. Springer Berlin
Heidelberg.
doi:\href{https://doi.org/10.1007/978-3-540-72446-9}{10.1007/978-3-540-72446-9}.

@@

\begin{figure}[htbp]
\centering
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{(a) Minimum time required to estimate change in abundance for
species class, (b) long-term trend (estimated slope coefficient) by
species class, (c) interannual variability in population size by species
class, and (d) temporal autocorrelation by species
class.\label{fig:class}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
corresponding fitted GAM model in red and (b) statistical power as a
function of the number of years sampled. The horizontal ine at 0.8
indicates the minimum threshold for statistical power and the vertical
line denotes the minimum time required to achieve 0.8 statistical
power.\label{fig:gam_example}}
\end{figure}

>>>>

To be submitted to: \vspace{3 mm}

@@

Long-term census data are necessary to better understand ecological
processes and make management decisions. However, census data are often
expensive, requiring a lot of people-hours and equipment. Little work
has addressed the question, when is a time series of census data long
enough to address a question of interest? Here, we explore two
approaches to address this question: simulations and an empirical
approach. We specifically determine the minimum time series length
required to estimate significant increases or decreases in population
abundance. Importantly, we examine the ability to detect a trend with a
set level of statistical power, which is often neglected in ecological
time series analyses. Using a simulation approach, we demonstrate how
the minimum time series length required increases with weaker trends in
abundance and with higher population variability. In addition, we
examine 868 populations of vertebrate species to determine the minimum
time required to detect changes in their abundance. We show that 10-15
years of continuous monitoring are required in order to achieve a high
level of statistical power. Similar to simulation-approaches, the
minimum time required for field census data strongly depends on trend
strength, population variability, and temporal autocorrelation. These
results point to the importance of sampling populations over long
periods of time. Most studies, especially those less than 10-15 years,
are probably under-powered and potentially misleading.

@@

Observational studies and population censuses have become a cornerstone
of modern ecological research and conservation biology (Magurran et al.
2010; Hughes et al. 2017). Long-term data are necessary to both
understand population dynamics and to assess extinction risk of species.
Even though many time series may now be considered ``long-term'', most
are still short. Time series are typically short for a variety of
reasons. They are often coupled with an experiment, which may only last
a couple of years. In addition, short funding cycles make it difficult
to examine populations over longer periods of time (Hughes et al. 2017).
There are, of course, notable exceptions where long-term monitoring has
occurred, for example the continuous plankton recorder (Giron-Nava et
al. 2017).

Naturally, one might ask, how long of a time series is actually
necessary? This question has important implications for both research
and management. Scientifically, we need to know the time series length
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when a trend in abundance over time is actually meaningful or
not. The IUCN Red List Categories and Criteria suggest under Criterion
A2, a species qualifies as vulnerable if it has experienced a 30\%
decline over 10 years, or 3 generations (IUCN 2012). For both scientific
and management questions, because sampling is typically expensive, we do
not want to sample for longer than is necessary (Gerber, DeMaster, and
Kareiva 1999).

@@

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (Bolker 2008). Instead of using classic statistical
tests (e.g.~t-test), and corresponding formulas for determining sample
size, simulation approaches can be used (Gerrodette 1987; Bolker 2008;
P. C. Johnson et al. 2015). Simple models can be simulated with
parameter values corresponding to a population of interest. Out of a
large number of simulations, power can be calculated as the fraction of
simulations that meet some criteria. The specific criteria depend on the
question at hand. Then, time series simulations of different lengths are
run to determine the minimum time series length required to be confident
that you will be able to answer your question (Bolker 2008). In addition
to using simulated data, empirical time series can also be used. We
usually do not have multiple replicates of similar populations, but we
can subsample an empirical time series (Gerber, DeMaster, and Kareiva
1999). Subsamples of different lengths can then be evaluated to see
which fraction of subsamples meet some criteria, again a measure of
statistical power. Similar to the simulation approach, we can use these
measures of power to determine the minimum time series required for a
particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized a model specifically for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organism, 15 years was adequate to assess long-term trends
in abundance. However, these studies, and other past work, have
typically focused either on theoretical aspects of monitoring design or
focused on single species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. More specifically, we estimate the minimum time
series length required to assess long-term changes in abundance via
simple linear regression. We examine 878 population time series to
estimate this minimum time series length required for detecting
population trends.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses (P. C.
Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
approach as it has been described elsewhere (citations). Essentially, we
take a stochastic population model and simulate it for a number of years
repetitively. We can then determine the minimum time series required to
achieve designed levels of significance and power for a particular
question. This approach requires us to determine values for our model
parameters (e.g.~growth rate). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
indicate that, if there was a true trend in abundance, we would have a
0.8 probability of detecting the trend.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) example of simulated time series for 40 time periods and
(b) statistical power versus the simulated time series length. The
horizontal, dashed line is our desired statistical power of 0.8. The
vertical, dashed line is the minimum time required to achieve our
desired statistical power. (c) Minimum time required (\(T_{min}\)) for
simulations with different values of the trend strength (\(r\)) and (d)
different levels of population variability (\(\sigma\)). In each case,
the minimum time required is the minimum number of years to achieve 0.8
statistical power given a significance level of
0.05.\label{fig:theoretical_approach}}
\end{figure}

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The database includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. Lastly, we also calculated the minimum number of years
required (10 years or 3 generations) under IUCN Red List Criterion A2 to
classify a species as vulnerable (IUCN 2012).

For a subset of populations in the database (\(n =\) 547), we were able
to connect a database on biological characteristics (Myhrvold et al.
2015) like body size and generation length. All 547 populations are
within the Aves class. In the supplementary material, we explore how the
minimum time required does not strongly correlate with any biological
characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 40
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then we
determined the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the long-term, or ``true'', time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1 - \beta\), where \(\beta\) is the
probability of a type II error). Lastly, we determine which subsample
length is required to guarantee we have above a certain threshold of
statistical power, which we will assume is 0.8 by convention (Cohen
1992).

@@

We also examined a subset of populations where more explanatory
variables were known (e.g.~body mass, longevity, location). None of
these explanatory variables were predictive of the minimum time series
length required (Fig. \ref{fig:biological_correlates}). Some differences
if \(T_{min}\) did occur between biological class (Fig.
\ref{fig:class}), but this can be attributed to the amount populations
are known to fluctuate (i.e.~the variability in population size).

@@

In the supplementary material, we explore several additional questions
related to statistical power and estimating \(T_{min}\). First, we
showed that other metrics besides estimating the slope coefficient from
linear regression could be used. More specifically, we show how the same
approach described here can be used to determine the minimum time
required to estimate a population's geometric growth rate (Figs.
\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored
how the simulation approach can be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
More specifically, we determine the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). We found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we
determine the minimum time required to estimate long-term trends
according to generalized additive models, instead of the simple linear
models used here (Fig. \ref{fig:gam_example}). We found that although we
obtain a similar distribution of minimum times required for GAMs, the
minimum time required for GAMs is on average shorter than for linear
regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these
results in the supplementary material.

@@

Here, we focus on an empirical approach to estimate the minimum time
series length required to assess changes in abundance over time. We
examined long time series and then subsampled each to determine the
minimum time required to achieve a desired level of significance and
power for linear regression. Statistical power is important as it
provides on information as to the necessary samples required to
determine a significant trend (Legg and Nagy 2006). We found that for
878 populations, at least 10-15 years of continuous monitoring were
necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line
with past work on a short-lived snail species (Rueda-Cediel et al. 2015)
and a long-lived whale species (Gerber, DeMaster, and Kareiva 1999). In
line with theoretical predictions (Rhodes and Jonzen 2011), we also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to our prior expectations, we also found
that \(T_{min}\) did not correlate with any biological variables of
interest (Fig. \ref{fig:biological_correlates}). We initially
hypothesized that species with longer lifespans or generation times may
require a longer sampling period. Our result could have been a result of
at least two effects. First, the database we used may not include a
diverse enough set of species with different life history traits.
Second, the question we poised, whether a population is increasing or
decreasing, was specifically concerned with trends in population density
over time. Therefore, life-history characteristics may be more important
for other questions (e.g.~extinction risk) where estimates minimum time
required are needed.

@@

Our work has some limitations in determining the minimum time series
length required. First, \(T_{min}\) is particular to the specific
question of interest. For example, in the supplementary material, we
used generalized additive models to detect population trends and came up
with a similar, but quantitatively different value for \(T_{min}\) (Fig.
\ref{fig:min_time_dist_gam}). An additional complication is that for our
empirical approach, the subsampling of the full time series allows for
estimates of power, but the individual subsamples are clearly not
independent of one another. Further, estimates of \(T_{min}\) depend on
chosen values of \(\alpha\) and \(\beta\) (Fig.
\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a
specific population model parameterized for each population of interest.
Then, model simulations could be used to estimate the minimum time
series required to address each specific question of interest. Clearly,
this is not always practical, especially if conducting analyses for a
wide array of species as we do here. In addition, our statistical models
suggest that \(T_{min}\) does not correlate with any biological
variables of interest, at least for the question of linear regression
(Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to
use these results to predict \(T_{min}\) for another population, even if
the population is of a species with a similar life-history to one in our
database.

@@

We use a database of 878 populations to determine the minimum time
series length required to determine population trends. This goes beyond
previous work that either focused on theoretical investigations or a
limited number of species. We show that to identify long-term changes in
abundance, 10-15 years of continuous monitoring are required (Fig.
\ref{fig:min_time_dist}). In line with theoretical predictions
(Gerrodette 1987), we also show that \(T_{min}\) is strongly correlated
with the long-term population trend (i.e.~rate of increase), variability
in population size, and the temporal autocorrelation (Fig.
\ref{fig:correlates}). Our work implies that for many populations,
census data less than 10-15 years are probably not reliable. This
stresses the importance of long-term monitoring programs. From both a
scientific and management perspective estimates of \(T_{min}\) are
important. If a time series is too short we lack statistical power to
reliably detect long-term population trends. In addition, a time series
that is too long may be a poor use of already limited funds (Gerber,
DeMaster, and Kareiva 1999). Future work should examine other species,
with a wider range of life history characteristics. In addition, similar
approaches can be used to determine the minimum time series length
required to address additional questions of interest.

@@

\hypertarget{ref-Cohen1992}{}
Cohen, Jacob. 1992. ``A power primer.'' \emph{Psychological Bulletin}
112 (1): 155--59.
doi:\href{https://doi.org/10.1037/0033-2909.112.1.155}{10.1037/0033-2909.112.1.155}.

@@

\hypertarget{ref-Johnson2015}{}
Johnson, Paul C, Sarah J Barry, Heather M Ferguson, and Pie Müller.
2015. ``Power analysis for generalized linear mixed models in ecology
and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

@@

\hypertarget{ref-Morrison2008}{}
Morrison, Shawm, and David S. Hik. 2008. ``When? Where? And for how
long? Census design considerations for an Alpine Lagomorph, the Collared
pika.'' In \emph{Lagomorph Biology}, 103--13. Springer Berlin
Heidelberg.
doi:\href{https://doi.org/10.1007/978-3-540-72446-9}{10.1007/978-3-540-72446-9}.

@@

\begin{figure}[htbp]
\centering
\caption{Minimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{(a) Minimum time required to estimate change in abundance for
species class, (b) long-term trend (estimated slope coefficient) by
species class, (c) inter-annual variability in population size by species
class, and (d) temporal autocorrelation by species
class.\label{fig:class}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
corresponding fitted GAM model in red and (b) statistical power as a
function of the number of years sampled. The horizontal line at 0.8
indicates the minimum threshold for statistical power and the vertical
line denotes the minimum time required to achieve 0.8 statistical
power.\label{fig:gam_example}}
\end{figure}
